The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Natural or tidal?

### 2025-09-24 thesis/introduction.tex
half-done discussion

With the exception of the Magellanic Clouds, most dwarf galaxy
satellites of the Milky Way (MW) are \emph{quenched}, with little to no
recent star formation \citep[e.g.,][]{weisz+2014}. Indeed, most faint MW
satellites contain old stellar populations which are \emph{relics} from
the early universe, consisting of many of the oldest and most metal poor
stars known. Understanding the properties of dwarf galaxies thus has
implications across astronomy, from cosmological structure formation on
small scales to the formation of metal-poor stellar populations.

@@

Dwarf galaxies have long raised conundrums for theories of galaxy
formation. The discovery of Fornax and Sculptor in 1938
\citep{shapley1938}\footnote{Technically, the Large and Small Magellanic
  Clouds (LMC, SMC) are also classified as dwarf galaxies, but these
  were likely always known to humans at southern latitudes.}, with no
known analogues, already presented such an enigma. Shapley presented
these dwarfs as a new type of \emph{stellar system} resembling the
Magellanic Clouds and globular clusters but did not attempt to speculate
on their exact nature. While dwarf galaxies were quickly understood to
be galaxies based on the inferred luminosities and sizes, their nature
remained unclear for decades
\citep[e.g.,][]{hodge1971, gallagher+wyse1994}.

The earliest spectroscopic work hinted that dwarf galaxies may contain
substantial amounts of dark matter. From early determinations of the
velocity dispersion for the Sculptor (Scl) and Ursa Minor (UMi) dwarf
spheroidal (dSph) galaxies, inferred mass-to-light ratios were at least
10 times larger than the values of globular clusters {[}GCs; e.g.,
\citet{aaronson1983}; \citet{aaronson+olszewski1987}{]}. While rather
uncertain initially, these values were later corroborated with larger
and more precise samples \citep[e.g.,][]{hargreaves+1994}. At the time,
several theories were proposed to explain these unusually high
mass-to-light ratios. Examples include: ongoing tidal disruption
inflating inferred velocity dispersions
\citep[e.g.,][]{kuhn+miller1989}, the presence of massive central black
holes \citep[e.g.,][]{strobel+lake1994}, or modified theories of gravity
\citep{milgrom1995}. Over time, however, a consensus developed where the
high mass-to-light ratios of dwarf galaxies was due to the presence of a
dark matter halo \citep[e.g.,][]{dekel+silk1986, wechsler+tinker2018}.
Since then, the properties of dwarf galaxies have played an increasingly
important role in our understanding of the clustering of dark matter on
small scales \citep[e.g.,][]{sales+2022, bullock+boylan-kolchin2017}.

Today, a common definition for a dwarf galaxy is a gravitationally bound
stellar system with dark matter \citep[or, more specifically, not
consistent with Newtonian dynamics of visible matter
alone,][]{willman+strader2012}. In contrast, star clusters (like GCs)
have no clear evidence for dark matter. The boundary between these two
classes becomes uncertain for faint, compact stellar associations, known
as ``ambiguous'' systems.

Dwarf galaxies span a large range of physical sizes, luminosities, and
morphologies. Broadly, there are three classes of dwarf galaxies based
on luminosity, as illustrated by Fig.~\ref{fig:galaxy_images}. Local
\textbf{bright dwarfs} with magnitudes \(-14 \gtrsim M_V \gtrsim  -18\)
or stellar masses
\(3\times10^7\,\Mo \lesssim M_\star \lesssim 10^9\,\Mo\) \footnote{assuming
  stellar mass-to-light ratio of 1, may be \(\sim 2\) for older
  populations}, often exhibit irregular morphologies and recent star
formation. Fig.~\ref{fig:galaxy_images} shows the Large Magellanic Cloud
(LMC) as an example of an irregular, bright dwarf galaxy, where most
stars are in a rotationally-supported thin disk (seen nearly face-on)
with a prominent bar. \textbf{Classical dwarf spheroidals} occupy
intermediate luminosities ( \(-7.7 \gtrsim M_V  \gtrsim -14\) or
\(10^5\,\Mo \lesssim M_\star \lesssim 3\times10^7\,\Mo\)). Typically,
these systems are old, non-star forming, gas-poor, and spheroidal. All
Milky Way satellites discovered before digital sky surveys are
classicals. The 12 classical dwarf satellites of our Galaxy are (in
order of decreasing luminosity) Sagittarius, Fornax, Leo I, Sculptor,
Antlia II, Leo II, Carina, Draco, Ursa Minor, Canes Venatici I, Sextans
I, and Crater II.\footnote{While formally the dwarf galaxy names we
  discuss contain ``dwarf spheroidal'' (dSph), e.g.~Sculptor dSph, we
  omit this suffix for brevity.} The \textbf{ultra-faint}s dwarfs (UFDs)
occupy the very faintest magnitude range (\(M_V \gtrsim -7.7\) or
\(M_\star \lesssim 10^5\,\Mo\)). These galaxies have minuscule stellar
masses, typically compact sizes, and very metal-poor stellar populations
\citep[see review][]{simon2019}. Altogether, known dwarf galaxies span
more than 15 orders in magnitude, or over 7 decades in stellar mass.

Of local dwarfs, the classical systems still remain among the most
studied galaxies with the best constrained parameters. These dwarfs also
contain large numbers of bright (giant) stars which can be followed up
spectroscopically \citep[e.g.,][]{tolstoy+2023, pace+2020}. As a result,
the determination of fundamental properties such as the position, size,
orientation, distances, proper motions, line-of-sight (LOS) velocities,
and velocity dispersions \(\sigma_\V\), are all relatively well
constrained today. \textbf{revisit this\ldots JFN cuts a similar
version}

Most well-studied dwarf galaxies lie near the Milky Way or in its
vicinity, the Local Group (LG) of galaxies. TheLG is defined as the
group consisting of galaxies within \(\sim 1\) Mpc from the Milky
Way-Andromeda centre \citep[e.g.,][ and references
therein]{mcconnachie2012}. Today, we know that the Milky Way system is
teeming with dwarfs, many of which are satellites of either the Milky
Way or(MW) or Andromeda (M31). Fig.~\ref{fig:mw_satellite_system} shows
the MW satellite system, including dwarf galaxies, globular clusters,
and ambiguous systems. Such a nearby population of dwarf galaxies are
useful for resolved, detailed studies aiming to understand the nature of
these systems.

@@

We only understand a fraction of the universe's composition. The leading
theory of cosmology, Lambda Cold Dark Matter (\LCDM{}), posits that the
universe is composed of about 68\% dark energy (\(\Lambda\)), 27\% dark
matter (DM), and 5\% regular baryons\footnote{\emph{Baryons} here means
  baryons+leptons, i.e.~any standard model massive fermion.}
\citep{planckcollaboration+2020}. While the composition of dark matter
and dark energy remains elusive, we know their general properties. Dark
energy drives the acceleration of the expansion of the universe on large
scales. We do not discuss dark energy here---it does not substantially
affect the Local Group today. Dark matter, instead, makes up the vast
majority of mass in galaxies. Typically, galaxies have baryonic to dark
matter ratios of between 1:5 to beyond 1:1000 for faint dwarf galaxies
\citep[e.g.,][]{hayashi+2023}.

@@

The very early universe was almost featureless. Our earliest
observations of the universe stem from the cosmic microwave background
(CMB)---revealing a uniform, isotropic, near-perfect blackbody emission.
But tiny perturbations in the CMB, temperature fluctuations of 1 part in
10,000, reveal the underlying seeds of large-scale cosmological
structure. In an expanding universe, gravitational instability makes CDM
overdensities grow and collapse hierarchically onto larger structures.
Initially, baryonic matter was coupled to radiation and resisted
collapse. Dark matter, only influenced by gravity instead, freely
collapsed into the first structures. For mass perturbations both smaller
than the horizon and overdense enough to collapse, these overdensities
of dark matter become self-gravitating, known as \emph{halos}. After
recombination, where electrons combined with atomic nuclei to form
atoms, baryons decoupled from radiation and fell into the dark matter
halos. The densest pockets of baryons later formed the first stars and
galaxies.

Dark matter halos, and their associated galaxies, rarely evolve in
isolation. Instead, structure formation is \emph{hierarchical}. Small
dark matter halos collapse first and hierarchically merge into
progressively larger halos
\citep[e.g.,][]{blumenthal+1984, white+rees1978, white+frenk1991}.
Hierarchical assembly is evident through the large scale structure of
the universe, remnants of past mergers within the Milky Way, and tidal
disruption of dwarf galaxies and their streams around nearby galaxies.

@@

In \LCDM{} cosmological simulations, dark matter halos are remarkably
self-similar. In \citet{NFW1996, NFW1997}, hereafter NFW, the authors
observe that the spherically-averaged density profiles \(\rho(r)\) are
universally well described by a two-parameter law,
\begin{equation}\protect\phantomsection\label{eq:nfw}{
\rho/\rho_s= \frac{1}{(r/r_s)(1+r/r_s)^2},
}\end{equation} where \(r_s\) is a scale radius and \(\rho_s\) a scale
density. This profile has shown remarkable success in describing \LCDM{}
halos across several orders of magnitude in mass. NFW profiles are
\emph{cuspy}, where the density rises like \(\rho \sim 1/r\) at small
radii \(r \ll r_s\). The steepness of the density profile increases
gradually with radius, and at large radii the density falls off like
\(\rho \sim 1/r^3\).

@@

The blue solid curve in Fig.~\ref{fig:nfw_density} shows an example NFW
halo for a galaxy with \(\vmax=40\,\kpc\) and \(\rmax=8\,\kpc\), or
\(M_{200} = 1\times10^10\,\Mo\) and \(c=12.5\), the cosmological mean
for a Fornax stellar mass galaxy.

The two parameters of an NFW profile are not independent. Lower-mass
dark matter halos often collapse earlier, when the universe was denser.
As a result, low mass subhalos tend to be more concentrated
\citep[e.g.,][]{NFW1997}. The relationship between \(M_{200}\) and c, or
the mass-concentration relation, describes the mean trend of
concentration with mass or, equivalently, the dependence of \(\vmax\) on
\(\rmax\) \citep[e.g.,][]{bullock+2001, ludlow+2016}. The left panel of
Fig.~\ref{fig:smhm} illustrates the present-day mass-concentration from
\citet{ludlow+2016}. While concentration tends to decrease with
increasing mass, the relation has substantial scatter. Other parameters
such as the halo spin or shape may affect the scatter of the
mass-concentration relation, but their effect is typically expected to
be small \citep{navarro+2010, dicintio+2013, dutton+maccio2014}.

\begin{figure}
\centering
\includegraphics[width=0.5\linewidth,height=\textheight,keepaspectratio]{figures/example_density_profiles.png}
\caption[Example density profiles]{Density profiles in log density
versus log radius for stars and dark matter in a Fornax-like galaxy. The
dark matter is more extended and massive than the star across the entire
galaxy.}\label{fig:nfw_density}
\end{figure}

@@

In \LCDM{} galaxy formation, the majority of mass in a dwarf galaxy
comes from the extended dark matter halo. Fig.~\ref{fig:nfw_density}
shows an example exponential stellar component for the Fornax dwarf
galaxy with the expected underlying dark matter halo. This galaxy has a
stellar mass \(M_\star \approx 2.5\times10^7\,\Mo\) with half-light
radius \(0.65\,\kpc\) \citep[ with \citet{woo+courteau+dekel2008}
mass-to-light ratio]{munoz+2018}, and a dark matter halo with
\(\vmax=40\,\kpc\) and \(\rmax=8\,\kpc\). Even when the stars are
densest (the inner regions), the dark matter remains nearly an order of
magnitude higher in density. Stars make a small contribution to the
gravitational structure of dwarf galaxies--indeed, stars are reasonably
approximated as tracer particles of the underlying dark matter halo. In
addition, the stellar component is confined to the central regions of
the dark matter halo.

Several factors affect the SMHM trend, including environment, assembly
history, tidal effects, and the details of galaxy formation. In
particular, effects like ram-pressure stripping (removal of gas in the
dwarf galaxy due to pressure from the host's circumgalactic medium) and
tidal removal of gas cause star formation to quench
\citep[e.g.,][]{christensen+2024}. Additionally, the time of formation
(relative to reionization) can influence the resulting stellar content
\citep{kim+2024}. Finally, tides influence both the dark matter and
stellar mass but in different amounts \citep[e.g.,][]{PNM2008}.
Consequently, tides may reduce the halo mass more than the stellar mass,
adding additional scatter to the SMHM trend, particularly for low-mass
satellites \citep[e.g.,][]{fattahi+2018}. Understanding the effects of
tides on Local Group dwarf galaxies is essential for determining where
and how these galaxies formed in a cosmological context.

\subsection{Open questions and ongoing challenges in dwarf
galaxies}\label{open-questions-and-ongoing-challenges-in-dwarf-galaxies}

Advances in observational techniques have accelerated progress in our
understanding of dwarf galaxies and introduced new questions. Deep
digital photometric surveys have more than quadrupled the number of
known Milky Way dwarf galaxies since 2005 \citep{simon2019}. Upcoming
surveys, like the Vera Rubin Observatory's Legacy Survey for Space and
Time \citep[LSST,][]{ivezic+2019}, will likely identify even fainter
dwarf galaxies. In addition, large aperture multi-object spectrographs
have revealed the detailed and complex inner chemodynamical structure of
dwarf galaxies \citep[e.g.,][]{battaglia+2006, fabrizio+2016}. Beyond
precise structural and kinematic properties of dwarf galaxies, modern
observations allow for the separation of multiple stellar populations,
detailed constraints on the dark matter density profiles, and hints of
tidal disruption or stellar halos. \textbf{does this ¶ (or section)
belong here?}

@@

However, these tensions have eased as a result of improved understanding
of baryonic physics. For example, recent hydrodynamic simulations in
particular have shown that strong feedback can produce dark matter cores
\citetext{\citealp[e.g.,][\citet{tollet+2016}]{navarro+eke+frenk1996}; \citealp{fitts+2017}; \citealp{benitez-llambay+2019}; \citealp{orkney+2021}}.
Altogether, the numerous past challenges for \LCDM{} in the dwarf galaxy
regime illustrate the opportunity for dwarf galaxies to test the
understanding of galaxy formation and dark matter physics.

@@

Determining absolute parallax measurement is facilitated by the
observation that stars in different regions of the sky are affected by
parallax motion with different phases. By imaging two regions separated
by 106.5 degrees on the same focal plane, \emph{Gaia} measures tiny
changes in relative positions of stars across small and large angles.
Combining measurements from multiple epochs across several years, an
absolute all-sky reference frame is derived from which parallax and
proper motions are derived. In addition to astrometry, \emph{Gaia}
measures photometry in the wide \emph{G} band (330-1050nm) and colours
from the blue photometer (BP, 330-680 nm) and red photometer (RP,
640-1050 nm). \emph{Gaia} additionally provides low resolution BP-RP
spectra and radial velocity measurements of bright stars \citep[of
magnitudes \(G_{\rm RVS} < 16\),][]{gaiacollaboration+2016}. For our
work, \emph{Gaia}'s most relevant measurements are \(G\) magnitude,
\(G_{\rm BP} - G_{\rm RP}\) colour, \((\alpha, \delta)\) position, and
\((\mu_{\alpha*}, \mu_\delta)\) proper motions.\footnote{The proper
  motions \(\mu_\alpha\) and \(\mu_\delta\) are the apparent rates of
  change in right ascension, \(\alpha\), and declination, \(\delta\),
  typically in units of mili-arcsecond (mas) per year.
  \(\mu_{\alpha*} = \mu_\alpha \cos \delta\) corrects for projection
  effects in \(\alpha\).}

@@

\emph{Gaia} has revolutionized our understanding of Milky Way structure.
For example, the 6D dynamical measurements and metallicities of MW stars
led to the (re)discovery of past mergers or Milky Way building blocks
like \emph{Gaia}-Sausage Enceladus
\citetext{\citealp[e.g.,][]{helmi+2018}; \citealp{belokurov+2018}; \citealp[but
see also][]{meza+2005}}, out-of-equilibrium structures like the
\emph{Gaia} snail \citep[e.g.,][]{antoja+2018}, and dynamical effects of
the Milky Way's spiral arms and the bar in the solar neighbourhood
\citep[ and references therein]{hunt+vasiliev2025}. Moving to the Milky
Way halo, \emph{Gaia} has helped find and constrain numerous stellar
streams \citep{ibata+malhan+martin2019, bonaca+price-whelan2025}.
Altogether, \emph{Gaia} has revealed the hierarchical formation and
complex, evolving structure of our own Galaxy.

For Milky Way satellites, \emph{Gaia} has improved orbital analysis and
facilitated stellar membership determination. Before \emph{Gaia}, few
galaxies had precisely measured proper motions, more recently from
Hubble Space Telescope observations \citep[e.g.,][]{sohn+2017}.
\emph{Gaia} allowed the first systematic and precise determinations of
Milky Way satellite proper motions \citep{pace+li2019, MV2020a}. While
the proper motion uncertainty of a typical dwarf member star is often
large, by combining the proper motions of 100s or 1000s of stars from
\emph{Gaia}, precise average proper motion measurements can be
determined, sometimes only limited by \emph{Gaia}'s systematic error
floor \citep[e.g.,][]{MV2020a}. Proper motions have thus ushered in a
new era for MW satellite dynamical studies, where we can derive precise
orbits for any satellite, assuming a given MW potential. In addition,
\emph{Gaia} helps establish membership by separating out contaminating
MW foreground stars. By measuring parallaxes and/or proper motions, many
background and foreground stars can be classified as non-members
\citep[e.g.,][]{battaglia+2022, jensen+2024}.

@@

Projected luminosity / stellar density profiles efficiently characterize
the radial structure of a galaxy. At its most basic, light profiles
synthesize properties such as the shape, location, size, and orientation
of a dwarf galaxy. In addition, the details of a stellar density profile
can help interpret a galaxy's assembly and dynamical history
\citep[e.g.,][]{penarrubia+2009, lee+2018, querci+2025}. We note that
for resolved galaxies, these profiles are in terms of stellar counts
instead of surface brightness.

@@

The \citet{king1962} profile, also an empirical fit to globular
clusters, is also used to describe dwarf galaxies, especially in older
literature. Using three parameters, a core radius \(R_c\), a truncation
radius \(R_t\), and a characteristic density, \(\Sigma_0\), the King
profile may be written as \begin{equation}{
\Sigma_{\rm K} = \Sigma_0\left(\frac{1}{\sqrt{1 + (R/R_c)^2}} - \frac{1}{\sqrt{1+(R_t/R_c)^2}}\right).
}\end{equation} In much of the older literature, \(R_t\) was interpreted
as a ``tidal radius'', after an analogous interpretation for globular
clusters \citep[e.g.,][]{hodge1961, IH1995}.

>>>>

With the exception of the Magellanic Clouds, most dwarf galaxy
satellites of the Milky Way (MW) are \emph{quenched}, with little to no
recent star formation \citep[e.g.,][]{weisz+2014}. Indeed, most faint MW
satellites contain old stellar populations which are \emph{relics} from
the early universe, consisting of many of the oldest and most metal poor
stars known \citep{simon2019}. Understanding the properties of dwarf
galaxies thus has implications across astronomy, from cosmological
structure formation on small scales to the formation of metal-poor
stellar populations.

@@

Dwarf galaxies have long raised conundrums for theories of galaxy
formation. The discovery of Fornax and Sculptor in 1938
\citep{shapley1938}\footnote{Technically, the Large and Small Magellanic
  Clouds (LMC, SMC) are also classified as dwarf galaxies, but these
  were likely always known to humans at southern latitudes.}, with no
known analogues, already presented such an enigma. H. Shapley presented
these dwarfs as a new type of \emph{stellar system} resembling the
Magellanic Clouds and globular clusters but did not attempt to speculate
on their exact nature. While dwarf galaxies were soon understood to be
galaxies based on the inferred luminosities and sizes, their nature
remained unclear for decades
\citep[e.g.,][]{hodge1971, gallagher+wyse1994}.

The earliest spectroscopic work hinted that dwarf galaxies may contain
substantial amounts of dark matter. From early velocity dispersion
measurements for dwarf spheroidal (dSph) galaxies, inferred
mass-to-light ratios were at least 10 times larger than for globular
clusters \citep[GCs, e.g.,][]{aaronson1983, aaronson+olszewski1987}.
While uncertain initially, these values were later corroborated with
larger and more precise samples \citep[e.g.,][]{hargreaves+1994}. At the
time, several theories were proposed to explain these unusually high
mass-to-light ratios. Examples include: ongoing tidal disruption
inflating inferred velocity dispersions
\citep[e.g.,][]{kuhn+miller1989}, the presence of massive central black
holes \citep[e.g.,][]{strobel+lake1994}, or modified theories of gravity
\citep{milgrom1995}. Over time, however, a consensus developed where the
high mass-to-light ratios of dwarf galaxies was due to the presence of a
dark matter halo \citep[e.g.,][]{dekel+silk1986, wechsler+tinker2018}.
Since then, the properties of dwarf galaxies have played an increasingly
important role in our understanding of the clustering of dark matter on
small scales \citep[e.g.,][]{sales+2022, bullock+boylan-kolchin2017}.

Today, a common definition for a dwarf galaxy is a gravitationally bound
stellar system with dark matter.\footnote{Or, more generally, systems
  inconsistent with Newtonian dynamics of visible matter alone,
  \citet{willman+strader2012}} In contrast, star clusters (like GCs)
have no clear evidence for dark matter. The boundary between these two
classes blurs for faint, compact stellar associations, known as
``ambiguous'' systems.

Dwarf galaxies span a large range of sizes, luminosities, and
morphologies. Broadly, there are three classes of dwarf galaxies based
on luminosity. Local \textbf{bright dwarfs} with magnitudes
\(-14 \gtrsim M_V \gtrsim  -18\) or stellar masses
\(3\times10^7\,\Mo \lesssim M_\star \lesssim 10^9\,\Mo\) \footnote{assuming
  stellar mass-to-light ratio of 1, may be \(\sim 2\) for older
  populations}, often exhibit irregular morphologies and recent star
formation. Fig.~\ref{fig:galaxy_images} shows the Large Magellanic Cloud
(LMC) as an example of an irregular, bright dwarf galaxy, where most
stars are in a rotationally-supported thin disk (seen nearly face-on)
with a prominent bar. \textbf{Classical dwarf spheroidals}\footnote{While
  formally the dwarf galaxy names we discuss contain ``dwarf
  spheroidal'' (dSph), e.g.~Sculptor dSph, we omit this suffix for
  brevity. Additionally, the 12 classical dwarf satellites of our Galaxy
  are (in order of decreasing luminosity) Sagittarius, Fornax, Leo I,
  Sculptor, Antlia II, Leo II, Carina, Draco, Ursa Minor, Canes Venatici
  I, Sextans I, and Crater II. Antlia II, Crater II, and Canes Venatici
  I are the only post digital sky surveys additions.} occupy
intermediate luminosities ( \(-7.7 \gtrsim M_V  \gtrsim -14\) or
\(10^5\,\Mo \lesssim M_\star \lesssim 3\times10^7\,\Mo\)). Typically,
these systems are old, non-star forming, gas-poor, and spheroidal. All
Milky Way satellites discovered before digital sky surveys are
classicals, and these systems remain among the best studied. The
\textbf{ultra-faint}s dwarfs (UFDs) occupy the very faintest magnitude
range (\(M_V \gtrsim -7.7\) or \(M_\star \lesssim 10^5\,\Mo\)). These
galaxies have minuscule stellar masses, typically compact sizes, and
very metal-poor stellar populations \citep[see review][]{simon2019}.
Altogether, known dwarf galaxies span more than 15 orders in magnitude,
or over 7 decades in stellar mass.

Most well-studied dwarf galaxies lie near the Milky Way or in its
vicinity, the Local Group (LG) of galaxies. The LG is defined as the
group consisting of galaxies within \(\sim 1\) Mpc from the Milky
Way-Andromeda centre \citep[e.g.,][ and references
therein]{mcconnachie2012}. Today, we know that the Milky Way system is
teeming with dwarfs, many of which are satellites of either the Milky
Way or(MW) or Andromeda (M31). Fig.~\ref{fig:mw_satellite_system} shows
the MW satellite system, including dwarf galaxies, globular clusters,
and ambiguous systems. Such a nearby population of dwarf galaxies are
useful for resolved, detailed studies aiming to understand the nature of
these systems.

@@

We only understand a fraction of the universe's composition. The leading
theory of cosmology, Lambda Cold Dark Matter (\LCDM{}), posits that the
universe is composed of about 68\% dark energy (\(\Lambda\)), 27\% dark
matter (DM), and 5\% regular old baryons\footnote{\emph{Baryons} here
  means baryons+leptons, i.e.~any standard model massive fermion.}
\citep{planckcollaboration+2020}. While the composition of dark matter
and dark energy remains elusive, we know their general properties. Dark
energy drives the acceleration of the expansion of the universe on large
scales. We do not discuss dark energy here---it does not substantially
affect the Local Group today. Dark matter, instead, makes up the vast
majority of mass in galaxies. Typically, galaxies have baryonic to dark
matter ratios of between 1:5 to beyond 1:1000 for faint dwarf galaxies
\citep[e.g.,][]{hayashi+2023}.

@@

The very early universe was almost featureless. Our earliest
observations of the universe stem from the cosmic microwave background
(CMB)---revealing a uniform, isotropic, near-perfect blackbody emission.
But tiny perturbations in the CMB, temperature fluctuations of 1 part in
10,000, reveal the underlying seeds of large-scale cosmological
structure. In an expanding universe, gravitational instability makes CDM
overdensities grow and collapse hierarchically onto larger structures.
Initially, baryonic matter was coupled to radiation and resisted
collapse. Dark matter, only influenced by gravity instead, freely
collapsed into the first structures. Mass perturbations sufficiently
small and overdense become self-gravitating structures, known as
\emph{halos}. After recombination, where electrons combined with atomic
nuclei to form atoms, baryons decoupled from radiation and fell into the
dark matter halos. The densest pockets of baryons later formed the first
stars and galaxies.

Dark matter halos, and their associated galaxies, rarely evolve in
isolation. Instead, \LCDM{} structure formation is \emph{hierarchical}.
Small dark matter halos collapse first and hierarchically merge into
progressively larger halos
\citep[e.g.,][]{blumenthal+1984, white+rees1978, white+frenk1991}.
Hierarchical assembly is evident through the large scale structure of
the universe, remnants of past mergers within the Milky Way, and tidal
disruption of dwarf galaxies and their streams around nearby galaxies.

@@

In \LCDM{} cosmological simulations, dark matter halos are remarkably
self-similar. In \citet{NFW1996, NFW1997}, hereafter NFW, the authors
observe that the spherically-averaged density profiles \(\rho(r)\) are
universally well described by a two-parameter law,
\begin{equation}\protect\phantomsection\label{eq:nfw}{
\rho/\rho_s= \frac{1}{(r/r_s)(1+r/r_s)^2},
}\end{equation} where \(r_s\) is a scale radius and \(\rho_s\) a scale
density. This profile has shown remarkable success in describing \LCDM{}
halos across several orders of magnitude in mass. NFW profiles are
\emph{cuspy}, where the density rises like \(\rho \sim 1/r\) at small
radii \(r \ll r_s\). The steepness of the density profile increases
gradually with radius, and at large radii the density falls off like
\(\rho \sim 1/r^3\). The blue solid curve in Fig.~\ref{fig:nfw_density}
shows an example NFW halo.

@@

The two parameters of an NFW profile are not independent. Lower-mass
dark matter halos often collapse earlier, when the universe was denser.
As a result, low mass subhalos tend to be more concentrated
\citep[e.g.,][]{NFW1997}. The relationship between \(M_{200}\) and c, or
the mass-concentration relation, describes the mean trend of
concentration with mass or, equivalently, the dependence of \(\vmax\) on
\(\rmax\) \citep[e.g.,][]{bullock+2001, ludlow+2016}. The left panel of
Fig.~\ref{fig:smhm} illustrates the present-day mass-concentration from
\citet{ludlow+2016}. While concentration tends to decrease with
increasing mass, the relation has substantial scatter. Other parameters
such as the halo spin or shape may affect the scatter of the
mass-concentration relation, but their effect is typically expected to
be small \citep{navarro+2010, dicintio+2013, dutton+maccio2014}. The
mass-concentration relation informs our assumed structure of Scl and UMi
later.

\begin{figure}
\centering
\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{figures/example_density_profiles.png}
\caption[Example density profiles]{Density profiles in log density
versus log radius for stars and dark matter in a Fornax-like galaxy. The
dark matter is more extended and massive than the star across the entire
galaxy. The halo has \(\vmax=40\,\kpc\) and \(\rmax=8\,\kpc\), or
\(M_{200} = 1\times10^{10}\,\Mo\) and \(c=12.5\), the cosmological mean
for a Fornax stellar mass galaxy. The galaxy has a stellar mass
\(M_\star \approx 2.5\times10^7\,\Mo\) with half-light radius
\(0.65\,\kpc\)
\citep{munoz+2018, woo+courteau+dekel2008}.}\label{fig:nfw_density}
\end{figure}

@@

In \LCDM{} galaxy formation, the majority of mass in a dwarf galaxy
comes from the extended dark matter halo. Fig.~\ref{fig:nfw_density}
shows an example exponential stellar component for the Fornax dwarf
galaxy with the expected underlying dark matter halo. Even when the
stars are densest in the inner regions, the dark matter remains nearly
an order of magnitude higher in density. Stars make a small contribution
to the gravitational structure of dwarf galaxies---indeed, stars are
reasonably approximated as tracer particles of the underlying dark
matter halo. In addition, the stellar component is confined to the
central regions of the dark matter halo.

Several factors affect the SMHM trend, including environment, assembly
history, tidal effects, and the details of galaxy formation. For
example, effects like ram-pressure stripping (removal of gas in the
dwarf galaxy due to pressure from the host's circumgalactic medium) and
tidal removal of gas cause star formation to quench
\citep[e.g.,][]{christensen+2024}. Additionally, the time of formation
(relative to reionization) can influence the resulting stellar content
\citep{kim+2024}. Finally, tides influence both the dark matter and
stellar mass but in different amounts, adding additional scatter to the
SMHM trend for satellites \citep[e.g.,][]{PNM2008, fattahi+2018}.
Understanding the effects of tides on Local Group dwarf galaxies aids in
revealing where and how these galaxies formed in a cosmological context.

\subsection{Challenges and questions concerning dwarf
galaxies}\label{challenges-and-questions-concerning-dwarf-galaxies}

@@

However, some of these tensions have eased as a result of improved
understanding of baryonic physics. For example, recent hydrodynamic
simulations in particular have shown that strong feedback can produce
dark matter cores
\citetext{\citealp[e.g.,][\citet{tollet+2016}]{navarro+eke+frenk1996}; \citealp{fitts+2017}; \citealp{benitez-llambay+2019}; \citealp{orkney+2021}}.
Several open questions remain, concerning e.g.~the plane of satellites,
and the details of the sizes and rotation curves of dwarf galaxies, the
existence and nature of stellar halos and accretion in dwarf galaxies
\citep[e.g.,][]{sales+2022}. Altogether, the numerous past and ongoing
challenges for \LCDM{} in the dwarf galaxy regime illustrate the
opportunity for dwarf galaxies to test the understanding of galaxy
formation and dark matter physics.

@@

Determining absolute parallax measurement is facilitated by the
observation that stars in different regions of the sky are affected by
parallax motion with different phases. By imaging two regions separated
by 106.5 degrees on the same focal plane, \emph{Gaia} measures changes
in relative positions of stars across small and large angles. Combining
measurements from multiple epochs across several years, an absolute
all-sky reference frame is derived from which parallax and proper
motions are derived. In addition to astrometry, \emph{Gaia} measures
photometry in the wide \emph{G} band (330--1050nm) and colours from the
blue photometer (BP, 330--680 nm) and red photometer (RP, 640--1050 nm).
\emph{Gaia} additionally provides low resolution BP-RP spectra and
radial velocity measurements of bright stars \citep[of magnitudes
\(G_{\rm RVS} < 16\),][]{gaiacollaboration+2016}. For our work,
\emph{Gaia}'s most relevant measurements are \(G\) magnitude,
\(G_{\rm BP} - G_{\rm RP}\) colour, \((\alpha, \delta)\) position, and
\((\mu_{\alpha*}, \mu_\delta)\) proper motions.\footnote{The proper
  motions \(\mu_\alpha\) and \(\mu_\delta\) are the apparent rates of
  change in right ascension, \(\alpha\), and declination, \(\delta\),
  typically in units of mili-arcsecond (mas) per year.
  \(\mu_{\alpha*} = \mu_\alpha \cos \delta\) corrects for projection
  effects in \(\alpha\).}

@@

\emph{Gaia} has revolutionized our understanding of Milky Way structure.
For example, the 6D dynamical measurements and metallicities of MW stars
led to the (re)discovery of past mergers or Milky Way building blocks
like \emph{Gaia}-Sausage Enceladus
\citetext{\citealp[e.g.,][]{helmi+2018}; \citealp{belokurov+2018}; \citealp[but
see also][]{meza+2005}}, out-of-equilibrium structures like the
\emph{Gaia} snail \citep[e.g.,][]{antoja+2018}, and dynamical effects of
the Milky Way's spiral arms and the bar in the solar neighbourhood
\citep[ and references therein]{hunt+vasiliev2025}. In the Milky Way
halo, \emph{Gaia} has helped find and constrain numerous stellar streams
\citep{ibata+malhan+martin2019, bonaca+price-whelan2025}. Altogether,
\emph{Gaia} has revealed the hierarchical formation and complex,
evolving structure of our own Galaxy.

For Milky Way satellites, \emph{Gaia} has improved orbital analysis and
facilitated robust stellar membership determination. Before \emph{Gaia},
few galaxies had precisely measured proper motions \citep[e.g.~using
Hubble Space Telescope,][]{sohn+2017}. \emph{Gaia} allowed the first
systematic and precise determinations of Milky Way satellite proper
motions \citep{pace+li2019, MV2020a}. While the proper motion
uncertainty of a typical dwarf member star is often large, by combining
the proper motions of 100s or 1000s of stars from \emph{Gaia}, precise
average proper motion measurements can be determined, sometimes only
limited by \emph{Gaia}'s systematic error floor
\citep[e.g.,][]{MV2020a}. Proper motions have thus ushered in a new era
for MW satellite dynamical studies, where we can derive precise orbits
for any satellite, assuming a given MW potential. In addition,
\emph{Gaia} helps establish membership by separating out contaminating
MW foreground stars. By measuring parallaxes and/or proper motions, many
more background and foreground stars can be classified as non-members
\citep[e.g.,][]{battaglia+2022, jensen+2024}.

@@

Projected luminosity / stellar density profiles efficiently characterize
the radial structure of a galaxy. At its most basic, light profiles
synthesize properties such as the shape, size, and orientation of a
dwarf galaxy. In addition, the details of a stellar density profile can
help interpret a galaxy's assembly and dynamical history
\citep[e.g.,][]{penarrubia+2009, lee+2018, querci+2025}. Note that for
resolved galaxies, these profiles are expressed in stellar count
densities instead of surface brightness.

@@

The \citet{king1962} profile, also a fit to globular clusters, is also
used to describe dwarf galaxies, more so in older literature. Using
three parameters, a core radius \(R_c\), a truncation radius \(R_t\),
and a characteristic density, \(\Sigma_0\), the King profile may be
written as \begin{equation}{
\Sigma_{\rm K} = \Sigma_0\left(\frac{1}{\sqrt{1 + (R/R_c)^2}} - \frac{1}{\sqrt{1+(R_t/R_c)^2}}\right).
}\end{equation} In much of the older literature, \(R_t\) was interpreted
as a ``tidal radius'', after an analogous interpretation for globular
clusters \citep[e.g.,][]{hodge1961, IH1995}.
